Kaa: A Python Implementation of Reachable Set Computation Using Bernstein Polynomials

### 2021-01-26 Complexity/LMNTheorem/paper.tex
Reorganized papers into more appropriate categories

\subsection{Characters}

We introduce some stepping stones defining the tools used in the Fourier Analysis of boolean functions. Many of these tools can be defined using techniques from Representation Theory of Finite Groups, but for the sake of briefity, we will not delve into those concepts here in detail. Instead, we will provide remarks and snippets on relevant topics along with references to some literature at the end of the section. Our presentation follows that of \cite{hatami}. \newline

@@

To define the Fourier transform, we must consider the Hilbert space $L_2(G)$ where $G$ endowed with the uniform probability measure i.e the discrete measure mapping each subset $H \subseteq G$ to $\frac{|H|}{|G|}$. These will be the complex-valued maps $f: G \rightarrow \mathbb{C}$ such that
\begin{equation}
  \frac{1}{|G|}\sum_{x \in G} |f(x)|^2 < \infty
\end{equation}
%
\noindent This yields the standard inner product for maps $f,g \in L_2(G)$: \newline
%
\begin{equation}
  \langle f,g \rangle = \frac{1}{|G|}\sum_{x\in G} f(x)\overline{g(x)}= \mathbb{E}_x f(x)\overline{g(x)}
\end{equation}
%
This induces the norm $||f||_2^2 = \langle f, f \rangle = \mathbb{E}_x |f(x)|^2$ on $L_2(G)$. It is simple to check that the characters of $G$ indeed lie in $L_2(G)$. However, we can prove something stronger: the characters of $G$, $\chi_a$ form an \emph{orthonormal basis} of $L_2(G)$.
%
First, we prove a lemma:
\begin{lemma} \label{totalsum}
  Let $\chi$ be a non-trivial character of $G$. Then $\sum_{x \in G} \chi(x) = 0$
\end{lemma}
%
\begin{proof}
  Given any non-trivial character $\chi$, the following holds for all $y \in G$: \newline
    %
    \[ \chi(y)\sum_{x \in G} \chi(x) = \sum_{x \in G} \chi(y+x) = \sum_{x \in G} \chi(x) \]\newline
    %
    If $\sum_{x \in G} \chi(x) \neq 0$, then this would force $\chi(y) = 1$ for all $y \in G$. However, this would contradict the non-triviality of the $\chi$.
\end{proof}

@@

\begin{remark}{\emph{(Generalizations of Fourier Analysis)}}
The above properties on characters can be shown through techniques in Representation Theory. For instance, we know representations of finite groups are group homomorphisms into the group of linear automorphisms of a suitable finite-dimensional complex vector space $\rho: G \rightarrow GL(\mathbb{C}^n)$. The \emph{characters} of $\rho$ are defined as $\chi(x) = Tr(\rho(x))$ where $Tr$ is the trace operator. It turns out that the representations of finite abelian groups are one-dimensional i.e isomorphic to $\mathbb{C}$. Thus, $\rho(x)$ will just be multiplying $\mathbb{C}$ by some scalar $\lambda_x \in \mathbb{C}^*$. In this case, we can show that indeed for any $a,b \in G$ and $\chi$ any character of $G$:
$$ \chi(a+b) = Tr(\rho(a+b)) = Tr(\rho(a)\circ\rho(b)) = \lambda_a \lambda_b = Tr(\rho(a))Tr(\rho(b)) = \chi(a)\chi(b)$$ Note that this relation does not hold for arbitrary finite groups. The genealizations of the Fourier Transform generally lie in \emph{Harmonic Analysis}. For more information on the Representation Theory of Finite Groups, see \cite{serre}, \cite{webb}.
\end{remark}

@@

  Since we assume that the $\AC^0$ circuit computing $f$ is in alternating normal form, once we invoke the switching lemma for the bottom two layer gates, we can merge the two $\vee$-gates to reduce the depth by one. The switching tells us that with probability of at least $1 - (5\frac{1}{10s}s)^s < 2^{-s}$, we can reduce the depth in each step such that the gates of distance two from the input variables have bottom fan-in of at most $s$. By the end of our depth-reduction routine, we are left with a CNF with bottom fan-in of at most $s$. Invoking Corollary (\ref{hastaddegree}) tells us that our final restricted function $f_{\rho_0\rho_1...\rho_{d-1}}$ has Fourier degree of at most $s$ with probability $< 2^{-s}$. The final bound arises from the observation that we applied the beginning analysis for each of the bottom gates exactly once, the switching lemma exactly once for each gate except for the output gate, and we invoked the corollary for the output gate. Each step contributed probability of $< 2^{-s}$.

@@

\begin{remark}{(\emph{Some concluding remarks})}
  he bounds shown by Linai-Mansour-Nisan have been improved by Tal (\cite{tal}). Specifically, he showed that for any $n$-ary boolean function $f$ with size $M$ and depth $d$,\
  \begin{equation}
    \sum_{|S| > t} |\hat{f}(S)|^2 = 2^{\Omega(k/\log^{d-1}{M})}
  \end{equation}
  Furthermore, he showed a converse to the proof direction employed by Linial et. al \cite{lmn}. Recall that in our proof we started with a switching lemma which bound the probability of the randomly-restricted function's Fourier degree exceeding a certain value. We then used this switching lemma to bound the tail Fourier spectrum. Tal showed a form of the opposite direction in \cite{talshrinkage}.
  \begin{theorem}{\emph{(Tal \cite{talshrinkage})}}
    Let $f\{1,-1\}^n \rightarrow \{-1,1\}$ be a Boolean function, let t,$C > 0$ such that for all $k$, $\sum_{|S| > k} |\hat{f}(S)|^2 \leq C e^{k/t}$ and let $\rho$ be a $p$-random restriction, then for all $d$, $Pr[\text{deg}(f_{\rho})] \leq C \cdot (4pt)^d$.
  \end{theorem}
  His work uses results in Quantum Query Complexity to ultimately culminate to another proof of the shrinkage exponent of Demorgan Formulae under random restrictions:
  \begin{theorem}{\emph{(Tal \cite{talshrinkage})}}
    Let $f$ be a Boolean function. For $p > 0$,
    \begin{equation}
      \mathbb{E}_{\rho \sim \mathcal{R}_p}[\mathcal{L}(f_\rho)] = \mathcal{O} \left(p^2\mathcal{L}(f) + p\sqrt{\mathcal{L}(f)} \right)
    \end{equation}
  \end{theorem}
\end{remark}
%    Bibliographies can be prepared with BibTeX using amsplain,
%    amsalpha, or (for "historical" overviews) natbib style.
\bibliographystyle{amsplain}
\bibliography{biblio}

>>>>

\subsection{Characters}
We introduce some stepping stones defining the tools used in the Fourier Analysis of boolean functions. Many of these tools can be defined using techniques from Representation Theory of Finite Groups, but for the sake of briefity, we will not delve into those concepts here in detail. Instead, we will provide remarks and snippets on relevant topics along with references to some literature at the end of the section. Our presentation follows that of \cite{hatami}. \newline

@@

To define the Fourier transform, we must consider the Hilbert space $L_2(G)$ where $G$ endowed with the uniform probability measure i.e the discrete measure mapping each subset $H \subseteq G$ to $\frac{|H|}{|G|}$. These will be the complex-valued maps $f: G \rightarrow \mathbb{C}$ such that
\begin{equation}
  \frac{1}{|G|}\sum_{x \in G} |f(x)|^2 < \infty
\end{equation}
%
\noindent which yields the standard inner product for maps $f,g \in L_2(G)$: \newline
%
\begin{equation}
  \langle f,g \rangle = \frac{1}{|G|}\sum_{x\in G} f(x)\overline{g(x)}= \mathbb{E}_x f(x)\overline{g(x)}
\end{equation}
%
This induces the norm $||f||_2^2 = \langle f, f \rangle = \mathbb{E}_x |f(x)|^2$ on $L_2(G)$. It is simple to check that the characters of $G$ indeed lie in $L_2(G)$. However, we can prove something stronger: the characters of $G$, $\chi_a$ form an \emph{orthonormal basis} of $L_2(G)$.
%
First, we prove a lemma:
\begin{lemma} \label{totalsum}
  Let $\chi$ be a non-trivial character of $G$. Then $\sum_{x \in G} \chi(x) = 0$
\end{lemma}
%
\begin{proof}
  Given any non-trivial character $\chi$, the following holds for all $y \in G$: \newline
    %
    \[ \chi(y)\sum_{x \in G} \chi(x) = \sum_{x \in G} \chi(y+x) = \sum_{x \in G} \chi(x) \]\newline
    %
    If $\sum_{x \in G} \chi(x) \neq 0$, then this would force $\chi(y) = 1$ for all $y \in G$. However, this would contradict the non-triviality of the $\chi$.
\end{proof}

@@

\begin{remark}{\emph{(Generalizations of Fourier Analysis)}}
The above properties on characters can be shown through techniques in Representation Theory. For instance, we know representations of finite groups are group homomorphisms into the group of linear automorphisms of a suitable finite-dimensional complex vector space $\rho: G \rightarrow GL(\mathbb{C}^n)$. The \emph{characters} of $\rho$ are defined as $\chi(x) = Tr(\rho(x))$ where $Tr$ is the trace operator. It turns out that the representations of finite abelian groups are one-dimensional i.e isomorphic to $\mathbb{C}$. Thus, $\rho(x)$ will just be multiplying $\mathbb{C}$ by some scalar $\lambda_x \in \mathbb{C}^*$. In this case, we can show that indeed for any $a,b \in G$ and $\chi$ any character of $G$:
$$ \chi(a+b) = Tr(\rho(a+b)) = Tr(\rho(a)\circ\rho(b)) = \lambda_a \lambda_b = Tr(\rho(a))Tr(\rho(b)) = \chi(a)\chi(b)$$ Note that this relation does not hold for arbitrary finite groups since the trace is generally not multiplicative. The genealizations of Fourier Analysis generally lie in \emph{Harmonic Analysis}. For more information on the Representation Theory of Finite Groups, see \cite{serre}, \cite{webb}.
\end{remark}

@@

  Since we assume that the $\AC^0$ circuit computing $f$ is in alternating normal form, once we invoke the switching lemma for the bottom two layer gates, we can merge the two $\vee$-gates to reduce the depth by one. The switching tells us that with probability of at least $1 - (5\frac{1}{10s}s)^s < 2^{-s}$, we can reduce the depth in each step such that the gates of distance two from the input variables have bottom fan-in of at most $s$. By the end of our depth-reduction routine, we are left with a CNF with bottom fan-in of at most $s$. Invoking Corollary (\ref{hastaddegree}) tells us that our final restricted function $f_{\rho_0\rho_1...\rho_{d-1}}$ has Fourier degree of at most $s$ with probability $< 2^{-s}$. The final bound arises from the observation that we applied the beginning analysis for each of the bottom gates exactly once, the switching lemma exactly once for each gate above the bottom gates except for the output gate, and we invoked the corollary for the output gate. Each step contributed probability $< 2^{-s}$.

@@

\begin{remark}{(\emph{Some concluding remarks})}
  The bounds shown by Linai-Mansour-Nisan have been improved by Tal (\cite{tal}). Specifically, he showed that for any $n$-ary boolean function $f$ with size $M$ and depth $d$
  \begin{equation}
    \sum_{|S| > t} |\hat{f}(S)|^2 = 2^{\Omega(k/\log^{d-1}{M})}
  \end{equation}
  Furthermore, he showed a converse to the proof direction employed by Linial et. al \cite{lmn}. Recall that in our proof we started with a switching lemma which bound the probability of the randomly-restricted function's Fourier degree exceeding a certain value. We then used this switching lemma to bound the tail Fourier spectrum. Tal showed a form of the opposite direction in \cite{talshrinkage}.
  \begin{theorem}{\emph{(Tal \cite{talshrinkage})}}
    Let $f\{1,-1\}^n \rightarrow \{-1,1\}$ be a Boolean function, let $t,C > 0$ such that for all $k$, $\sum_{|S| > k} |\hat{f}(S)|^2 \leq C e^{k/t}$ and let $\rho$ be a $p$-random restriction, then for all $d$, $Pr[\text{deg}(f_{\rho})] \leq C \cdot (4pt)^d$.
  \end{theorem}
  His work uses results in Quantum Query Complexity to ultimately culminate into another proof of the shrinkage exponent of Demorgan Formulae under random restrictions:
  \begin{theorem}{\emph{(Tal \cite{talshrinkage})}}
    Let $f$ be a Boolean function. For $p > 0$,
    \begin{equation}
      \mathbb{E}_{\rho \sim \mathcal{R}_p}[\mathcal{L}(f_\rho)] = \mathcal{O} \left(p^2\mathcal{L}(f) + p\sqrt{\mathcal{L}(f)} \right)
    \end{equation}
  \end{theorem}
\end{remark}
%    Bibliographies can be prepared with BibTeX using amsplain,
%    amsalpha, or (for "historical" overviews) natbib style.
\bibliographystyle{amsplain}
\bibliography{biblio}

### 2021-01-26 Nonlinear/KAA-Arch/samplepaper.tex
Reorganized papers into more appropriate categories

\begin{document}
%
\title{Kaa: A Python Implementation of Reachable Set Computation Using Bernstein Polynomials}
%
\titlerunning{Kaa: Python Reachability Tool}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Edward Kim
%\orcidID{0000-1111-2222-3333} 
\and
Parasara Sridhar Duggirala
%\orcidID{1111-2222-3333-4444}}
}
%
\authorrunning{Kim and Duggirala}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Department of Computer Science\\
University of North Carolina at Chapel Hill\\
\email{\{ehkim,psd\}@cs.unc.edu}
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reachable set computation is one of the widely used techniques for verification of safety properties of dynamical system. 
%
One of the simplest algorithms for computing reachable sets for discrete nonlinear systems uses parallelotope bundles and Bernstein polynomials.
%For discrete nonlinear systems, using Bernstein polynomials and parallelotope bundles for computing reachable set have been investigated in the literature.
%One of the reachable set computation techniques that has received considerable attention uses Bernstein polynomials and parallelotope bundles.
%
In this paper, we describe Kaa, a terse Python implementation of reachable set computation which leverages the widely used symbolic package sympy.
%
Additionally, we simplify the user interface with the tool and provide easy to use plotting utilities.
%
%We also provide an easy to use interfact for plotting the reachable sets.
%
We believe that our tool has pedagogical value given the simplicity of the implementation and its user-friendliness.
%
%In addition to the simplicity of the implementation, our tool, named Kaa, also achieves better performance on standard benchmarks.
\end{abstract}

@@

Reachable set computation is one of the 
%many 
important tools %available 
for verification of safety properties of dynamical and hybrid systems.
%
A simpler and easier-to-understand reachable set computation algorithm that utilizes Bernstein polynomials and parallelotopes has been presented in~\cite{dang2009image}.
%
Several cumulative improvements~\cite{dang2012reachability,sassi2012reachability,dang2014parameter,dreossi2016parallelotope,dreossi2017reachability} have been proposed to improve its accuracy and efficiency.
%
The tool SAPO~\cite{dreossi2017sapo} which implements these algorithms in C\texttt{++} is available publicly.
%
In this paper, we present a Python implementation of algorithm presented in~\cite{dreossi2017reachability}.

%
Our motivation for reimplementing the algorithms in Python is two-fold.
%
First, using standard libraries for symbolic manipulation in Sympy, the algorithm for reachability can be implemented very concisely.
%
In fact, the main algorithm for computing the reachable set has be implemented in only one file with $150$ lines of code.
%
The total code base for Kaa is roughly only $650$ lines of code.
%
%Additionally, the main algorithm for computing the reachable set is provided in just one file with $150$ lines of code.
%
In contrast, SAPO is more than $2000$ lines of C\texttt{++} code with memory and pointer management performed completely by the developer.
%
We believe that such a simple implementation can be used for teaching one of the main algorithms for reachability of nonlinear systems.
%
% Second, because of its implementation in C++, SAPO requires recompiling the code for every reachable set computation.
% %
Second, reimplementing the tool in Python makes user interaction significantly easy. 
%
In SAPO, one has to recompile the code along with the model file to generate the binary and the user has to execute the binary for computing the reachable set.
%
Additionally, visualizing the reachable set using SAPO is a two step process. In the first step, SAPO generates a MATLAB script that contains projections of reachable set for visualizing time or phase plots of the reachable set. In the second step, the user feeds the generated script into either MATLAB or Octave to illustrate the reachable set. 
%
In contrast, Kaa has very intuitive interface for computing and visualizing reachable set.
%
The models and the code for computing reachable set using Kaa are provided in Python and do not read any compilation. 
%
We also leverage the matplotlib library for visualizing the reachable set. 
%
Integrating these two, we provide an easy to use Jupyter notebook interface for computing and visualizing reachable set with Kaa for some of the standard benchmark models.
%
%Juypter notebooks are simple to create and known for their straightforward user interface. %
%we were able to design an engaging interactive tutorial and experimentation platform for visualzing reachable sets of non-linear systems.
%
%Therefore, the user has to 
%we believe that for reimplementing it in Python improves the quality of user interaction with the tool and removes the barrier to use the tools.
%

Bernstein polynomials are also an active area of research in the domain of global optimization~\cite{nataray2002algorithm,garloff2003bernstein,nataraj2007new}.
%
Several heuristics have been proposed for improving the performance of optimization using Bernstein polynomials~\cite{kodiak,smith2009fast,munoz2013formalization}. 
%
%Furthermore, these algorithms also improve the accuracy using sequential refinement techniques~\cite{}.
%
Having an accessible and flexible tool would make it more conducive to implement these heuristics and determine if they are helpful in the domain of reachability.
%
In light of these features and advantages, we believe Kaa can be used as a easy first step for introducing reachability in a pedagogical setting.

@@

As mentioned earlier, a parallelotope $P$ can also be represented as an affine transformation $T_{p}$ from $[0,1]^{n}$ to $P$. 
%
Therefore, upper bounds on the suprenum of a function $h$ over $P$ is equivalent to upper bound of $h \circ T_{p}$ over $[0,1]^{n}$.
%
A similar argument follows for the lower bound on infimum.
%

For the remainder of the document, we assume that by using functional composition and the Bernstein representation, we can compute the upper bound on supremum and lower bound on infimum of polynomial functions over parallelotopes.
%
We denote the procedures for calculating such upper and lower bounds for a polynomial $h$ over some parallelotope $P$ as $\mathsf{BernsteinUpper}(h,P)$ and $\mathsf{BernsteinLower}(h,P)$ respectively.

@@

One of the primary reasons for preferring Python as an implementation language for this algorithm is the presence of powerful, well-tested symbolic and matrix-computation libraries.
%
Python's \emph{numpy} libraries are popular matrix-computation libraries which allow higher-level manipulation of matrices. This gives us an avenue of overcoming the verbosity and the possibility of memory leaks inherent in implementing identical features in C++. We believe that overcoming these obstacles results in a more readable, more compact approach to parallelotope reachability suitable for graduate students and curious practitioners. In addition, these libraries are known for their extensive set of useful functionalities which allow us to avoid reimplementing many fundamental operations.

%
As we have seen before, there are two main sub-routines for computing the reachable set using Bernstein polynomials.
%
First, performing functional composition for computing the upper bound of a polynomial over a parallelotope.
%
Second, computing the Bernstein representation of a polynomial.
%
The library of \emph{sympy} has powerful symbolic manipulation tools which allow us to comfortably perform many sensitive symbolic function composition of polynomials. 
%
We use \emph{sympy}'s internal representation of polynomials to (a) perform functional composition and (b) compute the Bernstein representations of the resulting polynomial.
%
Additionally, the \emph{matplotlib} library has easy plotting facilities that we integrate into our tool for visualizing the reachable set. In particular, \emph{matplotlib} facilitates the ability to plot several reachable sets simultaneously. This will aid others in performing more complex bundle-transformation experiments in future.

%
Finally, Python has a rich set of multiprocessing libraries, namely \emph{multiprocessing} or \emph{mp} for short. In the future, we plan to exploit the parallelizable nature of bundle-transformtion computations using \emph{mp}'s ability to provide powerful multiprocessing features without much verbosity.

@@

\section{Evaluations}
We evaluate our tool on the benchmarks that are shipped along with SAPO. With each benchmark, the reachable sets of SAPO are juxtaposed with that of Kaa's to demonstrate the accuracy and quality of our reachable set. A comparison of time taken by Kaa and SAPO for each benchmark is provided in Table~\ref{tab:one}.

@@

\subsection{Lotka-Volterra Model}
\noindent We also test on the competetitive Lotka-Volterra models for predator-prey biological systems. Our instance is a 5-dimensional sysmtem governed as below:
\begin{align}
    \begin{split}
        x_{k+1} &= x_k + (x_k(1 - (x_k + \alpha y_k + \beta \ell_k))) \Delta \\
        y_{k+1} &= y_k + (y_k(1 - (y_k + \alpha z_k + \beta x_k))) \Delta \\
        z_{k+1} &= z_k + (z_k(1 - (z_k + \alpha h_k + \beta y_k))) \Delta \\
        h_{k+1} &= h_k + (h_k(1 - (h_k + \alpha \ell_k + \beta z_k))) \Delta \\
        \ell_{k+1} &= \ell_k + (\ell_k(1 - (\ell_k + \alpha x_k + \beta h_k))) \Delta
    \end{split}
\end{align}
where $\alpha,\beta$ are parameters set to $\alpha=0.85$ and $\beta=0.5$. We set $\Delta = 0.01$.
The relevant reachable sets are shown in Figure \ref{fig4}.

\subsection{Phosphoraley Model}
The Phosphoraley model describes a certain cellular regulatory system. It is captured by seven variables governed by the following dynamics:
\begin{align}
    \begin{split}
        x^1_{k+1} &= x^1_k + ( -\alpha x^1_k + \beta x^3_k x^4_k)\Delta \\
        x^2_{k+1} &= x^2_k + (  \alpha x^1_k - x^2_k)\Delta \\
        x^3_{k+1} &= x^3_k + ( x^2_k \beta x^3_k x^4_k)\Delta \\
        x^4_{k+1} &= x^4_k + ( \beta x^5_k x^6_k - \beta x^3_k x^4_k)\Delta \\
        x^5_{k+1} &= x^5_k + ( -\beta x^5_k x^6_k + \beta x^3_k x^4_k)\Delta \\
        x^6_{k+1} &= x^6_k + ( \alpha x^7_k - \beta x^5_k x^6_k)\Delta \\
        x^7_{k+1} &= x^7_k + ( -\alpha x^7_k + \beta x^5_k x^6_k)\Delta
    \end{split}
\end{align}
where $\alpha,\beta$ are two parameters assigned as $\alpha=0.5$ and $\beta=5$. We set $\Delta =0.01$ here. The relevant figures can be found in Figure \ref{fig5}.

@@

\noindent It is evident from Table~\ref{tab:one} that while the current implementation in Python is very intuitive and concise, it incurs severe performance penalties. 
%
Nevertheless, we pursued this activity because of the pedagogical value in making this an accessible tool for implementing and understanding reachability. To this end, the repository contains a Jupyter notebook with the name \href{https://github.com/Tarheel-Formal-Methods/kaa/blob/master/kaa-intro.ipynb}{kaa-intro} designed to interactively introduce graduate students and practitioners to the usage of Kaa. The notebook even contains a subset of the examples shown above, giving the auidence a chance to run the code for themselves. An immediate next step is to deploy extensive profiling to find performance bottlenecks and subsequently improve on them.

@@

We present Kaa, a Python implementation of reachable set computation of nonlinear systems which is focused towards accessibility and pedagogical use.
%
The usage of inbuilt \emph{sympy} libraries makes the implementation short and simple (only 650 LOC).
%
While we do incur performance drawbacks from selecting Python for implementing this algorithm, we believe that it aids in fast prototyping and enables students to easily build on top of the library. In particular, Python's readability and extendibility will allow curious students to experiment with more involved bundle-transformations. 
%
The code can be accessed through \url{https://github.com/Tarheel-Formal-Methods/kaa}
\newline
\newline
{\bf Acknowledgements: } This material is based upon work supported by the Air Force Office of Scientific Research under award number FA9550-19-1-0288 and National Science Foundation (NSF) under grant numbers CNS 1739936, 1935724. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the United States Air Force or National Science Foundation.

>>>>

\begin{document}
%
\title{Kaa: A Python Implementation of Reachable Set Computation Using Bernstein Polynomials}
%
\titlerunning{Kaa: Python Reachability Tool}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Edward Kim
%\orcidID{0000-1111-2222-3333} 
\and
Parasara Sridhar Duggirala
%\orcidID{1111-2222-3333-4444}}
}
%
\authorrunning{Kim and Duggirala}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Department of Computer Science\\
University of North Carolina at Chapel Hill\\
\email{\{ehkim,psd\}@cs.unc.edu}
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Reachable set computation is one of the many widely-used techniques for the verification of safety properties of dynamical systems. 
%
One of the simplest algorithms for computing reachable sets for discrete nonlinear systems uses parallelotope bundles and Bernstein polynomials.
%For discrete nonlinear systems, using Bernstein polynomials and parallelotope bundles for computing reachable set have been investigated in the literature.
%One of the reachable set computation techniques that has received considerable attention uses Bernstein polynomials and parallelotope bundles.
%
In this paper, we describe Kaa, a terse Python implementation of reachable set computation which leverages the widely used symbolic package \emph{sympy}.
%
Additionally, we simplify the user interface and provide easy-to-use plotting utilities.
%
%We also provide an easy to use interfact for plotting the reachable sets.
%
We believe that our tool has pedagogical value given the simplicity of the implementation and its user-friendliness.
%
%In addition to the simplicity of the implementation, our tool, named Kaa, also achieves better performance on standard benchmarks.
\end{abstract}

@@

Reachable set computation is one of the 
%many 
important tools %available 
for verification of safety properties of dynamical and hybrid systems.
%
A simpler and easier-to-understand reachable set computation algorithm that utilizes Bernstein polynomials and parallelotopes has been presented in~\cite{dang2009image}.
%
Several cumulative improvements~\cite{dang2012reachability,sassi2012reachability,dang2014parameter,dreossi2016parallelotope,dreossi2017reachability} have been proposed to improve its accuracy and efficiency.
%
The tool SAPO~\cite{dreossi2017sapo} which implements these algorithms in C\texttt{++} is available publicly.
%
In this paper, we present a Python implementation of algorithm presented in~\cite{dreossi2017reachability}.
%
Our motivation for reimplementing the algorithms in Python is two-fold.
%
First, using standard libraries for symbolic manipulation in Sympy, the algorithm for reachability can be implemented concisely.
%
In fact, the main algorithm for computing the reachable set has be implemented in only one file with $150$ lines of code.
%
The total code base for Kaa is roughly only $650$ lines of code.
%
%Additionally, the main algorithm for computing the reachable set is provided in just one file with $150$ lines of code.
%
In contrast, SAPO consists of more than $2000$ lines of C\texttt{++} code with memory and pointer management performed completely by the developer.
%
We believe that such a simple implementation can be used for teaching one of the main algorithms for reachability of nonlinear systems.
%
% Second, because of its implementation in C++, SAPO requires recompiling the code for every reachable set computation.
% %
Second, reimplementing the tool in Python makes user interaction significantly simpler. 
%
In SAPO, one has to recompile the code along with the model file to generate the binary. Subsequently the user has to execute the binary to compute the reachable set.
%
Additionally, visualizing the reachable set using SAPO is a two step process. In the first step, SAPO generates a MATLAB script which plots projections of reachable set for visualizing time or phase plots of the reachable set. In the second step, the user feeds the generated script into either MATLAB or GNU Octave to illustrate the reachable set. 
%
In contrast, Kaa offers plotting functionalities through the \emph{matplotlib} library for visualizing the reachable set.
%
The models and the code for computing reachable set using Kaa are provided in Python and do not require any compilation steps.
%
Integrating these two, we provide an easy-to-use Jupyter notebook interface for computing and visualizing reachable set with Kaa for some of the standard benchmark models.
%
%Juypter notebooks are simple to create and known for their straightforward user interface. %
%we were able to design an engaging interactive tutorial and experimentation platform for visualzing reachable sets of non-linear systems.
%
%Therefore, the user has to 
%we believe that for reimplementing it in Python improves the quality of user interaction with the tool and removes the barrier to use the tools.
%

Bernstein polynomials are also an active area of research in the domain of global optimization~\cite{nataray2002algorithm,garloff2003bernstein,nataraj2007new}.
%
Several heuristics have been proposed for improving the performance of optimization using Bernstein polynomials~\cite{kodiak,smith2009fast,munoz2013formalization}. 
%
%Furthermore, these algorithms also improve the accuracy using sequential refinement techniques~\cite{}.
%
Having an accessible and flexible tool would make it more conducive to implement these heuristics and determine if they are helpful in the domain of reachability.
%
In light of these features and advantages, we believe Kaa could be used as a natural first step for introducing reachability in a pedagogical setting.

@@

As mentioned earlier, a parallelotope $P$ can also be represented as an affine transformation $T_{p}$ from $[0,1]^{n}$ to $P$. 
%
Therefore, upper bounds on the suprenum of a function $h$ over $P$ is equivalent to upper bound of $h \circ T_{p}$ over $[0,1]^{n}$.
%
A similar argument follows for the lower bound on the infimum.
%

For the remainder of the document, we assume that by using functional composition and the Bernstein representation, we can compute the upper bound on supremum and the lower bound on the infimum of polynomial functions over parallelotopes.
%
We denote the procedures for calculating such upper and lower bounds for a polynomial $h$ over some parallelotope $P$ as $\mathsf{BernsteinUpper}(h,P)$ and $\mathsf{BernsteinLower}(h,P)$ respectively.

@@

One of the primary reasons for preferring Python as the implementation language for this algorithm is the presence of powerful, well-tested symbolic and matrix-computation libraries.
%
Python's \emph{numpy} libraries are popular matrix-computation libraries which allow higher-level manipulation of matrices. This gives us an avenue of overcoming the verbosity and the possibility of memory leaks inherent in implementing identical features in C++. We believe that overcoming these obstacles results in a more readable, more compact approach to parallelotope reachability suitable for graduate students and curious practitioners. In addition, these libraries are known for their extensive set of useful functionalities which allow us to avoid reimplementing many fundamental operations.

%
As we have seen before, there are two main sub-routines for computing the reachable set using Bernstein polynomials.
%
First, performing functional composition for computing the upper bound of a polynomial over a parallelotope.
%
Second, computing the Bernstein representation of a polynomial.
%
The library of \emph{sympy} has powerful symbolic manipulation tools which allow us to comfortably perform many sensitive symbolic function composition of polynomials.
%
We use \emph{sympy}'s internal representation of polynomials to (a) perform functional composition and (b) compute the Bernstein representations of the resulting polynomial.
%
Additionally, the \emph{matplotlib} library has accessible plotting facilities that we integrate into our tool for visualizing the reachable set. In particular, \emph{matplotlib} facilitates the ability to plot several reachable sets simultaneously. This will aid others in performing more complex bundle-transformation experiments in the future.

%
Finally, Python contains a rich set of multiprocessing libraries, namely \emph{multiprocessing} or \emph{mp} for short. We plan to exploit the parallelizable nature of bundle-transformtion computations using \emph{mp}'s ability to provide powerful multiprocessing features without much verbosity.

@@

\section{Evaluations}
We evaluate our tool on the benchmarks that are shipped along with SAPO. With each benchmark, the reachable sets of SAPO are juxtaposed with that of Kaa's to demonstrate the accuracy and quality of our reachable sets. A comparison of time taken by Kaa and SAPO for each benchmark is provided in Table~\ref{tab:one}.

@@

\subsection{Lotka-Volterra Model}
\noindent We also test on the competetitive Lotka-Volterra models for predator-prey biological systems. Our instance is a 5-dimensional system governed as below:
\begin{align}
    \begin{split}
        x_{k+1} &= x_k + (x_k(1 - (x_k + \alpha y_k + \beta \ell_k))) \Delta \\
        y_{k+1} &= y_k + (y_k(1 - (y_k + \alpha z_k + \beta x_k))) \Delta \\
        z_{k+1} &= z_k + (z_k(1 - (z_k + \alpha h_k + \beta y_k))) \Delta \\
        h_{k+1} &= h_k + (h_k(1 - (h_k + \alpha \ell_k + \beta z_k))) \Delta \\
        \ell_{k+1} &= \ell_k + (\ell_k(1 - (\ell_k + \alpha x_k + \beta h_k))) \Delta
    \end{split}
\end{align}
where $\alpha,\beta$ are parameters set to $\alpha=0.85$ and $\beta=0.5$. We set $\Delta = 0.01$.
The relevant reachable sets are shown in Figure \ref{fig4}.

\subsection{Phosphoraley Model}
The Phosphoraley model describes a certain cellular regulatory system. It is captured by seven variables governed by the following dynamics:
\begin{align}
    \begin{split}
        x^1_{k+1} &= x^1_k + ( -\alpha x^1_k + \beta x^3_k x^4_k)\Delta \\
        x^2_{k+1} &= x^2_k + (  \alpha x^1_k - x^2_k)\Delta \\
        x^3_{k+1} &= x^3_k + ( x^2_k \beta x^3_k x^4_k)\Delta \\
        x^4_{k+1} &= x^4_k + ( \beta x^5_k x^6_k - \beta x^3_k x^4_k)\Delta \\
        x^5_{k+1} &= x^5_k + ( -\beta x^5_k x^6_k + \beta x^3_k x^4_k)\Delta \\
        x^6_{k+1} &= x^6_k + ( \alpha x^7_k - \beta x^5_k x^6_k)\Delta \\
        x^7_{k+1} &= x^7_k + ( -\alpha x^7_k + \beta x^5_k x^6_k)\Delta
    \end{split}
\end{align}
where $\alpha,\beta$ are two parameters assigned as $\alpha=0.5$ and $\beta=5$. We set $\Delta =0.01$ here. The figures can be found in Figure \ref{fig5}.

@@

\noindent It is evident from Table~\ref{tab:one} that while the current implementation in Python is intuitive and concise, it incurs severe performance penalties. 
%
Nevertheless, we pursued this direction because of the pedagogical value in making this an accessible tool for implementing and understanding reachability. To this end, the repository contains a Jupyter notebook with the name \href{https://github.com/Tarheel-Formal-Methods/kaa/blob/master/kaa-intro.ipynb}{kaa-intro} designed to interactively introduce graduate students and practitioners to the usage of Kaa. The notebook even contains a subset of the examples shown above, giving the auidence a hands-on method to run the code for themselves. An immediate next step is to deploy extensive profiling to find performance bottlenecks and subsequently improve on them.

@@

We present Kaa, a Python implementation of reachable set computation of nonlinear systems which is focused towards accessibility and pedagogical use.
%
The usage of inbuilt \emph{sympy} libraries makes the implementation short and simple (only 650 LOC).
%
While we do incur performance drawbacks from selecting Python for implementing this algorithm, we believe that it aids in fast prototyping and enables students to easily build on top of the library. In particular, Python's readability and extendibility will allow curious students to experiment with more involved bundle-transformations. 
%
The code can be accessed through \url{https://github.com/Tarheel-Formal-Methods/kaa}.
\newline
\newline
{\bf Acknowledgements: } This material is based upon work supported by the Air Force Office of Scientific Research under award number FA9550-19-1-0288 and National Science Foundation (NSF) under grant numbers CNS 1739936, 1935724. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the United States Air Force or National Science Foundation.
